\SW: Crowd-sourcing the Discovery of Gravitational Lenses

### 2014-05-13 doc/sw-system.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Space Warps I: Experiment Design
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title[\sw I]
{\SW: Crowd-sourcing the Discovery of Gravitational Lenses}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\date{to be submitted to MNRAS}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}\pubyear{2013}

@@

Scientific motivation. Applications of lenses: group-scale arcs,
galaxy-galaxy lenses, lensed quasars.

Problem of rarity. Imaging surveys. Problem of purity/false positives.

Review of progress to date. Methods in SL2S, SQLS. Contrast with SLACS.

Scaling to wide field era. Automated methods: problems. Need for good
training sets. Need for quality control: always present.

Novel solution: crowd-sourcing. Brief review of similar problems.
PlanetHunters. \sw as an experiment.

In this paper, we describe the \sw website, an online system that enables
crowd-sourced detection of gravitational lenses.  In a companion paper we will
present the new gravitational lenses discovered in our first imaging survey
dataset, and begin to investigate the differences between lens detections made
in \sw and those made with automated techniques. Here though, we try to answer
the following questions:

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Space Warps System Paper
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\title[\sw]
{\SW: Crowd-sourcing the Discovery of Gravitational Lenses}

@@

@@

\date{to be submitted to MNRAS}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}\pubyear{2014}

@@

% Scientific motivation. Applications of lenses: group-scale arcs,
% galaxy-galaxy lenses, lensed quasars.

Strong gravitational lensing -- the formation of multiple, magnified images of
background objects due to the deflection of light by  massive foreground
objects -- is a very powerful astrophysical tool, enabling a wide range of
science projects. The image separations and distortions provide information
about the mass distribution in the lens \citep[e.g.][]{Auger++2011,
Sonnenfeld++2014}, including on sub-galactic scales
\citep[e.g.][]{Dalal+Kochanek2004,Vegetti+Koopmans2009,Hezaveh++2012}. Any
strong lens can provide magnification of a factor of 10 or more, providing a
deeper, higher resolution view of the distant universe through these ``cosmic
telescopes'' \citep[e.g.][]{Stark++2009,Newton++2009}. Lensed quasars enable
cosmography via the time delays between the multiple images' lightcurves
\citep[e.g.][]{Tewes++2011,Suyu++2011}, and study of the accretion disk itself
through the microlensing effect \citep[e.g.][]{Morgan++2011}. All of these
science projects would benefit from being able to draw from a larger sample of
lenses.

In the last decade the numbers of detections of these rare cosmic alignments
has increased by an order of magnitude, thanks to wide field surveys such as
CLASS \citep[e.g.]{Browne++2003}, SDSS \citep[e.g.][]
{Bolton++2006,Auger++2010,Treu++2011,Oguri++2010}, CFHTLS
\citep[e.g.][]{More++2012,Gavazzi++2014}, Herschel \citep[][]{Herschel} and
SPT \citep[e.g.][]{Vieira++2012}, among others.  As the number of known lenses
has increased, new types have been discovered, leading to entirely new
investigations. Compound lenses \citep{Gavazzi++2007,Collett+Auger2014} and
lensed supernovae \citep{Quimby++2014} are good examples of this.

% Problem of rarity. Imaging surveys. Problem of purity/false positives.
% Review of progress to date. Methods in SL2S, SQLS. Contrast with SLACS.

Because they are rare, strong lenses are expensive to find. The most efficient
searches to date have made use of relatively clean signals such as the
presence of emission or absorption features at two distinct redshifts in the
same optical spectrum \citep[e.g.][]{Bolton++2004}, or the strong
``magnification bias'' towards detecting strongly-lensed sources in the sub-mm
waveband \citep[e.g.][]{Negrello++2010}. Such searches have to be efficient,
because they require expensive high resolution imaging follow-up; consequently
they have so far produced yields in the tens to hundreds. An alternative approach
is to search images of sufficiently high resolution and color contrast, and
confirm the systems as gravitational lenses by modeling the survey data
themselves \citep[][]{Marshall++2009}. Several square degrees of HST images
have been searched, yielding several tens of galaxy-scale lenses
\citep[e.g.][]{Moustakas++2006,Faure++2009,Jackson++2010,Pawase++2012}.
Similarly, searches of over a hundred square degrees of CFHT Legacy Survey
ground-based imaging, also with sub-arcsecond image quality, have revealed a
smaller number of wider image separation group-scale systems
\citep[e.g.][]{Cabanac++2007,More++2010}. Detecting galaxy-scale lenses from
the ground is hard, but feasible albeit lower efficiency and requiring HST or
spectroscopic follow-up to confirm the candidates as lenses
\citep[e.g.][]{Gavazzi++2014}.

% Scaling to wide field era. Automated methods: problems. Need for good
% training sets. Need for quality control: always present.

How can we scale these lens searches up to imaging surveys covering a hundred
times the sky area, such as the almost-all sky surveys planned with LSST and
Euclid, while reducing our dependence on expensive follow-up confirmation
observations? There are two approaches to detecting lenses in imaging surveys.
The first one is robotic: automated analysis of object catalogs and/or the
survey images. The candidate samples produced by these methods have, to date,
not been of high purity  \citep[see
e.g.][]{Marshall++2009,More++2010,Gavazzi++2014}, with visual inspection by
teams of humans still required to narrow down the robotically-generated
samples. In this approach, the image data may or may not be explicitly
modelled by the robots as if it contained a gravitational lens, but the visual
inspection can be thought of as a ``mental modeling'' step. Systems classified
by an inspector to be good lens candidates are deemed as such because the
features in the image can be explained by a model of what gravitational lenses
do contained in the inspector's brain. The second approach simply cuts out the
robot middleman: \citet{Faure++2009,Jackson++2010} and \citet{Pawase++2012}
all performed entirely visual searches for lenses in HST imaging.

Visual image inspection seems, at present, unavoidable at some level when
seraching for gravitational lenses. The technique has some drawbacks, however.
First is that humans are only humans, and they make mistakes. The solution to
this is to operate in teams, providing multiple classifications of the same
images in order to catch errors and correct them. Second, and relatedly, is
that humans get tired. With a well-designed classification interface, a human
might be able to inspect images at a rate of one astronomical object per
second (provided the majority are indeed uninteresting). At $10^4$ massive
galaxies, and 10 lenses, per square degree, visual lens searches in good
quality imaging data are limited to a few square degrees per inspector per
day. Scaling to thousands of square degrees therefore means either robotically
reducing the number of targets for inspection, or increasing the number of
inspectors, or both.

For example, a $10^4$ square degree survey containing $10^8$
photometrically-selected massive galaxies and $10^5$ lenses could only be
searched by 10 inspectors at a mean rate of 1 galaxy per second and 10
inspections per galaxy in about 14 years. Reducing the inspection time by a
factor of 400 to two weeks would require a robot to reduce the target sample
to 25 per square degree. However, at this point the required purity, 40\%,
would very likely require the average classification time per object to be
more like 10 seconds per object. Hiring 10 inspectors to assess complex images
full time full time for five months may not be the most cost-effective or
reliable strategy. Alternatively, a team of $10^6$ inspectors could, in
principle, make the required $10^9$ image classifications, $10^3$ each, in a
few weeks; robotically reducing the target list would lead to a proportional
decrease in the required team size.

Systematic detection of rare astronomical objects by such ``crowd-sourced''
visual inspection has recently been achieved by the online citizen science
project PlanetHunters \citep{PlanetHunters}. PlanetHunters was designed to
enable the discovery of transiting exoplanets in data taken by the Kepler
satellite; a community of N inspectors found, after each undergoing a small
amount of training, N new exoplanet candidates by visual inspection of the
Kepler lightcurves that were presented in a custom web-based classification
interface. The older Galaxy Zoo morphological classification project
\citep{GalaxyZoo} has also enabled the discovery of rare objects, via its
flexible inspection interface and discussion forum. Indeed, several of us
(AV,EB,CC,TJ,CM,LW) were active in an informal Galaxy Zoo gravitational lens
search, an experience which led to the present hypothesis that a systematic
online visual lens search could be successful.

In this paper, we describe the \sw website, an online system that enables 
crowd-sourced gravitational lens detection.  In a companion paper we will
present the new gravitational lenses discovered in our first experimental lens
search, and begin to investigate the differences between lens detections made
in \sw and those made with automated techniques. Here though, we try to answer
the following questions:
